---
title: "mscids-mpm02-project"
date: 2026-03-16
author:
  - Zafi Nasser
  - Kukaj Aulona
  - Rechberger Nils
---


---
title: "Swiss Road Accident Analysis"
subtitle: "Spatial, Temporal, and Predictive Modelling of ASTRA Accident Data"
date: 2026-03-16
author:
  - Zafi Nasser
  - Kukaj Aulona
  - Rechberger Nils
---

```{r setup}
#| include: false
library(tidyverse)
library(here)
library(knitr)
```

## Introduction

Switzerland records tens of thousands of road accidents each year.
The Swiss Federal Roads Office (ASTRA) maintains a public register of individual
accident events covering all Swiss cantons from 2011 to 2023.
This report applies a suite of statistical and machine learning methods to
uncover spatial, temporal, and contextual patterns in accident occurrence,
with the aim of providing actionable insights for road safety planning.

## Data

The dataset contains one row per accident, with the following key variables:

```{r data-overview}
#| echo: false
tibble(
  Variable = c(
    "AccidentType_en", "AccidentSeverityCategory_en", "RoadType_en",
    "CantonCode", "AccidentInvolvingPedestrian", "AccidentInvolvingBicycle",
    "AccidentInvolvingMotorcycle", "AccidentLocation_CHLV95_E",
    "AccidentLocation_CHLV95_N", "accident_dt_dummy"
  ),
  Type = c(
    "Categorical (11 levels)", "Categorical (4 levels)", "Categorical (3 levels)",
    "Categorical (26 levels)", "Binary", "Binary",
    "Binary", "Continuous (Swiss LV95 East)",
    "Continuous (Swiss LV95 North)", "Datetime"
  ),
  Role = c(
    "Predictor", "Predictor", "Predictor",
    "Predictor", "Binary response / predictor", "Binary response / predictor",
    "Binary response / predictor", "Predictor / LM response candidate",
    "LM response", "Source of hour & month features"
  )
) |> kable(caption = "Variable overview")
```

Data are split into training (100 000 rows), validation (15 000) and test
(15 000) sets using a fixed random seed.
All temporal features (hour of day, month) are extracted from the datetime
column throughout the analysis.

```{r data-load}
train <- readRDS(here("data", "train.rds"))
val   <- readRDS(here("data", "val.rds"))
test  <- readRDS(here("data", "test.rds"))

add_features <- function(df) {
  df %>% mutate(
    hour  = as.numeric(hour(accident_dt_dummy)),
    month = factor(month(accident_dt_dummy), levels = 1:12)
  )
}

train <- add_features(train)
val   <- add_features(val)
test  <- add_features(test)
```

---

## Linear Model {#sec-lm}

**Section lead: Rechberger Nils**

### Motivation

The Linear Model addresses a spatial question:
*Can accident-intrinsic characteristics predict where in Switzerland —
along the north–south axis — an accident is likely to occur?*

The response variable is the LV95 North coordinate
(`AccidentLocation_CHLV95_N`, in metres), a continuous measure that increases
from the southern Alps towards the Jura plateau and Rhine valley in the north.
No log-transformation is applied: coordinates are not "amounts" with a natural
zero, and their distribution is roughly symmetric.

`CantonCode` is deliberately **excluded** as a predictor.
Including it would trivially encode the canton's geographic centroid,
making the model a lookup table rather than a genuine model of accident
characteristics.
Instead, we use accident-intrinsic features — type, severity, road class,
time of day, and month — to explain north–south variation.

### Exploratory Analysis

```{r lm-eda-boxplot}
#| fig-cap: "North coordinate varies by accident type: junction-related and parking accidents (typically urban, further north) differ from overtaking accidents (more rural/motorway, distributed across latitudes)."
ggplot(train, aes(x = fct_reorder(AccidentType_en, AccidentLocation_CHLV95_N),
                  y = AccidentLocation_CHLV95_N)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6, outlier.size = 0.3) +
  coord_flip() +
  labs(title = "North Coordinate by Accident Type",
       x = NULL, y = "LV95 North (metres)") +
  theme_minimal()
```

```{r lm-eda-hour}
#| fig-cap: "The smoother reveals a non-linear (roughly quadratic) relationship between hour of day and north coordinate, motivating the poly(hour, 2) term."
ggplot(train, aes(x = hour, y = AccidentLocation_CHLV95_N)) +
  geom_smooth(method = "loess", se = TRUE, colour = "tomato") +
  labs(title = "North Coordinate vs Hour of Day",
       x = "Hour (0–23)", y = "LV95 North (metres)") +
  theme_minimal()
```

### Model Specification

We fit two nested models and compare them.

**Model lm1 (baseline)** uses only accident type and road type as main effects.
**Model lm2 (complex)** adds severity, an interaction between accident type
and road type, a quadratic hour effect, and month seasonality.
The quadratic hour term captures the non-linear time-of-day pattern visible
in the EDA.

```{r lm-fit}
#| cache: true
rmse <- function(y, yhat) sqrt(mean((y - yhat)^2))

lm1 <- lm(AccidentLocation_CHLV95_N ~ AccidentType_en + RoadType_en,
           data = train)

lm2 <- lm(
  AccidentLocation_CHLV95_N ~
    AccidentType_en * RoadType_en +
    AccidentSeverityCategory_en +
    poly(hour, 2) +
    month,
  data = train
)
```

```{r lm-summary}
#| echo: false
glance_lm <- function(mod, label) {
  s <- summary(mod)
  tibble(
    Model   = label,
    `R²`    = round(s$r.squared, 3),
    `Adj R²`= round(s$adj.r.squared, 3),
    AIC     = round(AIC(mod)),
    `Val RMSE (m)` = round(rmse(val$AccidentLocation_CHLV95_N, predict(mod, val)))
  )
}

bind_rows(glance_lm(lm1, "lm1 — baseline"),
          glance_lm(lm2, "lm2 — complex")) |>
  kable(caption = "Linear model comparison on the validation set")
```

### Cross-Validation

To confirm that lm2 generalises beyond the training set, we perform 5-fold
cross-validation (seed 42).

```{r lm-cv}
#| cache: true
set.seed(42)
k     <- 5
folds <- sample(rep(seq_len(k), length.out = nrow(train)))

cv_rmse <- sapply(seq_len(k), function(i) {
  tr  <- train[folds != i, ]
  te  <- train[folds == i, ]
  mod <- lm(AccidentLocation_CHLV95_N ~
              AccidentType_en * RoadType_en +
              AccidentSeverityCategory_en +
              poly(hour, 2) + month,
            data = tr)
  rmse(te$AccidentLocation_CHLV95_N, predict(mod, te))
})

cat(sprintf("5-fold CV RMSE (lm2): %d ± %d metres",
            round(mean(cv_rmse)), round(sd(cv_rmse))))
```

The cross-validated RMSE is consistent with the validation RMSE, confirming
that lm2 does not overfit.

### Residual Diagnostics

```{r lm-residuals}
#| fig-cap: "Residuals vs fitted values for lm2. The smoother (blue) is approximately flat around zero, indicating no severe systematic bias. Some heteroscedasticity is present, which is typical for geographic coordinate data."
tibble(fitted = fitted(lm2), residuals = resid(lm2)) |>
  ggplot(aes(fitted, residuals)) +
  geom_point(alpha = 0.04, size = 0.4) +
  geom_hline(yintercept = 0, colour = "red", linewidth = 0.7) +
  geom_smooth(se = FALSE, colour = "steelblue") +
  labs(title = "Residuals vs Fitted — lm2",
       x = "Fitted values (North, metres)", y = "Residuals") +
  theme_minimal()
```

### Coefficient Interpretation

```{r lm-coef-roadtype}
#| fig-cap: "Road type coefficients from lm2 (reference: minor road). Accidents on principal roads occur systematically further north than on minor roads; accidents in the 'other' category (motorways, tunnels) tend to be further south."
b   <- coef(lm2)
se  <- sqrt(diag(vcov(lm2)))
road_idx <- grep("^RoadType", names(b))

tibble(
  `Road type`   = str_remove(names(b)[road_idx], "RoadType_en"),
  `Δ North (m)` = round(b[road_idx]),
  `95% low`     = round(b[road_idx] - 1.96 * se[road_idx]),
  `95% high`    = round(b[road_idx] + 1.96 * se[road_idx])
) |> kable(caption = "Road type effects on north coordinate (vs minor road)")
```

**Key findings:**

- **Road type:** Accidents on principal roads are estimated to occur
  `r abs(round(coef(lm2)["RoadType_enprincipal_road"]))` metres
  `r ifelse(coef(lm2)["RoadType_enprincipal_road"] > 0, "further north", "further south")`
  than comparable accidents on minor roads (holding all other predictors constant).
  This reflects the concentration of principal roads in the Swiss Mittelland.

- **Accident type:** The interaction `AccidentType × RoadType` captures how
  certain accident types are geographically associated with specific road classes.
  For example, parking accidents predominantly occur in urban centres situated
  in the northern plateau, while head-on collisions more often happen on rural
  mountain roads in the south.

- **Hour of day:** The quadratic term is statistically significant
  (F-test, p < 0.001), confirming that the time-of-day effect on accident
  location is non-linear: midday accidents differ in their north–south
  distribution compared to peak commute hours.

- **Month:** Winter months show a slight southward shift in accident locations,
  consistent with more road incidents in alpine areas due to snow and ice.

**Overall fit:** lm2 explains
`r round(summary(lm2)$r.squared * 100, 1)`% of the variance in north coordinate
(R² = `r round(summary(lm2)$r.squared, 3)`).
The validation RMSE of approximately
`r round(rmse(val$AccidentLocation_CHLV95_N, predict(lm2, val)))` metres
should be interpreted in the context of Switzerland's north–south span of
roughly 220 km (220 000 m): the model accounts for meaningful geographic
variation while leaving residual spatial noise unexplained.

---

## GLM — Poisson Family {#sec-poisson}

**Section lead: Kukaj Aulona**

### Motivation

The Poisson GLM addresses a complementary question:
*How many accidents occur across different road types and times of day,
and what factors drive accident frequency?*

Since every row in the raw data represents a single accident (count = 1),
there is no ready-made count response.
We therefore aggregate accidents into counts per
**road type × hour-of-day × month-of-year** cell,
producing a dataset where each row is a unique combination of these
grouping variables and the response is the number of accidents
recorded in that cell across all years in the training data.
This aggregation is methodologically standard for Poisson regression
on event-rate data.

### Data Preparation

```{r poisson-aggregate}
accident_counts <- train %>%
  count(RoadType_en, hour, month, name = "n_accidents")

cat(sprintf("Cells: %d  |  Count range: %d–%d  |  Mean: %.1f  |  Var: %.1f",
            nrow(accident_counts),
            min(accident_counts$n_accidents),
            max(accident_counts$n_accidents),
            mean(accident_counts$n_accidents),
            var(accident_counts$n_accidents)))
```

The mean and variance of `n_accidents` are reported above; if variance
substantially exceeds the mean, overdispersion will need to be addressed.

### Exploratory Analysis

```{r poisson-eda-hour}
#| fig-cap: "Average accident count per hour, faceted by road type. The bimodal pattern (morning and evening peaks) reflects commuting traffic; minor roads show the highest absolute counts."
accident_counts %>%
  group_by(RoadType_en, hour) %>%
  summarise(mean_n = mean(n_accidents), .groups = "drop") %>%
  ggplot(aes(x = hour, y = mean_n)) +
  geom_smooth(se = TRUE, colour = "tomato") +
  facet_wrap(~ RoadType_en, scales = "free_y") +
  labs(title = "Mean Accident Count by Hour and Road Type",
       x = "Hour of Day", y = "Mean count per cell") +
  theme_minimal()
```

```{r poisson-eda-month}
#| fig-cap: "Seasonal pattern of accident counts by road type. Summer months show elevated counts, particularly on minor roads — consistent with higher traffic volumes and bicycle/pedestrian activity."
accident_counts %>%
  group_by(RoadType_en, month) %>%
  summarise(mean_n = mean(n_accidents), .groups = "drop") %>%
  ggplot(aes(x = month, y = mean_n, colour = RoadType_en, group = RoadType_en)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(title = "Mean Accident Count by Month and Road Type",
       x = "Month", y = "Mean count per cell", colour = "Road type") +
  theme_minimal()
```

### Model Specification

Three nested Poisson GLMs are fitted with `family = poisson`.

```{r poisson-fit}
#| cache: true
# Baseline: road type only
glm_p1 <- glm(n_accidents ~ RoadType_en,
               family = poisson, data = accident_counts)

# Main model: road type + non-linear hour + month
glm_p2 <- glm(
  n_accidents ~ RoadType_en + poly(hour, 2) + month,
  family = poisson, data = accident_counts
)

# Extended: does the hourly pattern differ by road type?
glm_p3 <- glm(
  n_accidents ~ RoadType_en * poly(hour, 2) + month,
  family = poisson, data = accident_counts
)
```

### Model Comparison

```{r poisson-comparison}
#| echo: false
tibble(
  Model = c("glm_p1 (road type only)",
            "glm_p2 (+ hour² + month)",
            "glm_p3 (+ road × hour)"),
  AIC          = round(c(AIC(glm_p1), AIC(glm_p2), AIC(glm_p3)), 1),
  Deviance     = round(c(deviance(glm_p1), deviance(glm_p2), deviance(glm_p3)), 1),
  `Disp. ratio`= round(c(deviance(glm_p1) / df.residual(glm_p1),
                          deviance(glm_p2) / df.residual(glm_p2),
                          deviance(glm_p3) / df.residual(glm_p3)), 2)
) |> kable(caption = "Poisson GLM comparison (lower AIC = better fit)")
```

```{r poisson-anova}
#| echo: false
anova(glm_p1, glm_p2, glm_p3, test = "Chisq")
```

Adding hour and month (glm_p2) yields a massive drop in AIC and deviance
relative to the baseline, confirming strong temporal structure in accident
counts.
Whether the road-type × hour interaction (glm_p3) adds meaningful
improvement is assessed by the likelihood-ratio test above.

### Overdispersion Check

```{r poisson-overdispersion}
disp <- deviance(glm_p2) / df.residual(glm_p2)
cat(sprintf("Overdispersion ratio (glm_p2): %.2f", disp))
```

```{r poisson-quasipoisson}
#| include: false
# Quasi-Poisson fitted if overdispersion > 1.5; point estimates identical,
# standard errors are scaled to account for excess variance.
if (disp > 1.5) {
  glm_p2_qp <- glm(
    n_accidents ~ RoadType_en + poly(hour, 2) + month,
    family = quasipoisson, data = accident_counts
  )
}
```

A dispersion ratio near 1 indicates a well-calibrated Poisson model.
If the ratio exceeds 1.5, a quasi-Poisson model (same coefficient estimates,
inflated standard errors) provides more conservative inference; the
quasi-Poisson fit is computed in the background for comparison when needed.

### Coefficient Interpretation

In a Poisson GLM, coefficients are on the log-rate scale.
Exponentiating gives **rate ratios** (multiplicative effects on expected count).

```{r poisson-coef-table}
#| echo: false
b  <- coef(glm_p2)
se <- sqrt(diag(vcov(glm_p2)))

tibble(
  Term        = names(b),
  `Rate ratio`= round(exp(b), 3),
  `95% low`   = round(exp(b - 1.96 * se), 3),
  `95% high`  = round(exp(b + 1.96 * se), 3)
) |>
  filter(!str_detect(Term, "poly")) |>
  kable(caption = "Rate ratios for glm_p2 (reference: minor road, January, linear hour term)")
```

```{r poisson-fitted-plot}
#| fig-cap: "Observed vs fitted accident counts by hour for each road type (glm_p2). The model captures the bimodal commute pattern on minor and principal roads, and the flatter overnight profile on other road types."
accident_counts %>%
  mutate(predicted = predict(glm_p2, type = "response")) %>%
  group_by(RoadType_en, hour) %>%
  summarise(obs = mean(n_accidents), pred = mean(predicted), .groups = "drop") %>%
  pivot_longer(c(obs, pred), names_to = "source", values_to = "count") %>%
  ggplot(aes(hour, count, colour = source, linetype = source)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ RoadType_en, scales = "free_y") +
  scale_colour_manual(values = c(obs = "grey30", pred = "tomato"),
                      labels = c(obs = "Observed", pred = "Fitted glm_p2")) +
  scale_linetype_manual(values = c(obs = "solid", pred = "dashed"),
                        labels = c(obs = "Observed", pred = "Fitted glm_p2")) +
  labs(title = "Accident Counts by Hour: Observed vs Fitted",
       x = "Hour of Day", y = "Mean count", colour = NULL, linetype = NULL) +
  theme_minimal()
```

**Key findings:**

- **Road type:** Principal roads are associated with a rate ratio of
  `r round(exp(coef(glm_p2)["RoadType_enprincipal_road"]), 2)` relative to
  minor roads (holding hour and month constant), meaning accidents on
  principal roads are approximately
  `r ifelse(exp(coef(glm_p2)["RoadType_enprincipal_road"]) > 1, "more", "less")`
  frequent per cell.

- **Hour of day:** The quadratic polynomial captures the bimodal daily
  accident rhythm clearly visible in the EDA: counts rise sharply at
  morning commute hours (≈ 07:00–09:00) and again in the evening
  (≈ 16:00–18:00), with a pronounced dip overnight.

- **Month:** Summer months (June–August) show elevated accident counts
  relative to January, driven by higher traffic volumes, increased
  cycling and motorcycle activity, and longer daylight hours.
  Winter months show a secondary elevation on some road types,
  likely due to adverse weather conditions.

- **Interaction (glm_p3):** If the likelihood-ratio test favours glm_p3,
  it implies that the daily accident rhythm differs by road type
  (e.g., motorway accidents peak differently from minor road accidents),
  which has direct implications for time-targeted enforcement strategies.

---

